---
title: "Aggregating a general-to-general MPM"
author:
  - name: "Richard A. Hinrichsen"
    orcid: "0000-0003-0761-3005"
output:
  rmarkdown::html_vignette:
    number_sections: false
    mathjax: default
vignette: >
  %\VignetteIndexEntry{Aggregating a general-to-general MPM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
**ORCID:** <https://orcid.org/0000-0003-0761-3005>

## Overview

Matrix population models (MPMs) often differ in the number and definition of life-cycle 
stages used to represent population structure, making comparisons difficult. This is 
particularly common for size- or stage-structured models, where fine-grained stage 
definitions may be aggregated to facilitate analysis, interpretation, 
or comparison of MPMs.

In this vignette, we demonstrate how to aggregate a general MPM 
to another general MPM using the `mpmaggregate` package. Unlike 
aggregation approaches that modify the projection interval, general-to-general MPM
aggregation reconciles differences in stage resolution while leaving the projection
interval unchanged. Multiple fine-grained stages are combined into a smaller 
set of biologically meaningful stages, yielding a simplified representation of the life 
cycle.

As a worked example, we use a size-structured matrix population model for *Rourea induta* 
originally published by Hoffmann (1999) and obtained from the COMPADRE database 
(Salguero-Gómez et al., 2015), accessed via `Rcompadre` (Jones et al., 2022). 
The model contains multiple fine-grained size stages and is therefore well suited 
to illustrate how aggregation can be used to simplify stage structure while preserving 
key demographic properties (Salguero-Gómez & Plotkin, 2010).

Throughout the vignette, we emphasize interpretation: how aggregation alters the 
representation of the life cycle, what demographic information is preserved, and how 
the resulting aggregated model can be used for comparison or subsequent demographic 
analyses, such as estimating population growth rates, generation time, or elasticities. 
Although *Rourea induta* is used here as an example, the approach is broadly applicable 
to general stage-structured matrix population models across taxa.

In addition, we emphasize how different aggregators influence the reduced model. 
The vignette compares all four alternative aggregation approaches of `mpm_aggregate()`, 
which differ in the demographic quantities they are designed to preserve and in the 
criteria used to fit aggregated transitions. By examining similarities and differences 
among these approaches, we illustrate how methodological choices affect the structure 
and interpretation of aggregated stage-structured models.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Packages
We begin by loading the packages used to access matrix population models, perform 
matrix aggregation, and format output for display.
```{r packages}
library(mpmaggregate)
library(Rcompadre)
library(kableExtra)
```

## Data acquisition
To demonstrate general-to-general aggregation, we use a 14 × 14 matrix population 
model for *Rourea induta* (MatrixID 246781) from the COMPADRE database (Hoffmann 1999),
which has a projection interval of 1 year.

```{r fetch-comadre}
compadre <- cdb_fetch("compadre")
iii <- (compadre$Genus == "Rourea") & (compadre$Species == "induta")
compadre.rourea.induta<- compadre[iii, ]

# Development note: save metadata for future use
#write.table(cdb_metadata(compadre.rourea.induta),file="C:/Users/Rich/projects/mpmaggregate/inst/extdata/rourea-induta.csv",sep=",", col.names=NA)

#return the matrix A from compadre.rourea.induta
#associated with MatrixID matid
get_matrix <- function(matid) {
  compadre2 <- compadre.rourea.induta[compadre.rourea.induta$MatrixID == matid, ]
   matA <- matA(compadre2)[[1]];
   matU <- matU(compadre2)[[1]];
   matF <- matF(compadre2)[[1]]
   matC <- matC(compadre2)[[1]];
   return(list(matA=matA,matU=matU,matF=matF,matC=matC))}

mats <- get_matrix(246781)
matA <- mats$matA
matU <- mats$matU
matF <- mats$matF
matC <- mats$matC
matF <- matF + matC

#sanity check
stopifnot(all.equal(unname(matA), unname(matU + matF)))

#1                active          Seedling
#2                active            Sucker
#3                active         1-1 .9 mm
#4                active         2-3 .9 mm
#5                active         4-5 .9 mm
#6                active         6-7 .9 mm
#7                active        8-.9 .9 mm
#8                active       10-14 .9 mm
#9                active        15-19.9 mm
#10               active          20-29 mm
#11               active          30-39 mm
#12               active          40-49 mm
#13               active          50-69 mm
#14               active            70+ mm

#form the aggregated groups, leaving Seedling and Sucker stages alone
groups <- list(
  c(1),                #Seedling
  c(2),                #Sucker
  c(3, 4, 5, 6),       #1-7.9 mm
  c(7, 8, 9, 10),      #8-29 mm
  c( 11, 12, 13, 14)   #30+ mm
)
```
## Helper functions for demographic quantities

We compute:

- \(\lambda\): spectral radius of \( \mathbf{A} \)
- \(R_0\): spectral radius of \( \mathbf{F} \left( \mathbf{I} - \mathbf{U} \right)^{-1} \)
- \(T_a\): generation time (years) under the \(\lambda\) framework
- \(T_c\): cohort generation time (years) under the \(R_0\) framework

```{r helpers}
`%||%` <- function(x, y) if (!is.null(x)) x else y

get_R0 <- function(U, F) {
  I <- diag(nrow(U))
  N <- solve(I - U)
  K <- F %*% N
  spectral_radius(K)
}

get_Ta <- function(U, F) {
  generation_time(F, U, framework = "lambda")$generation_time
}

get_Tc <- function(U, F) {
  generation_time(F, U, framework = "R0")$generation_time
}
```

## Aggregate in four ways

We aggregate the model using the following four combinations of framework and criterion:

1. `framework = "lambda"`, `criterion = "standard"`
2. `framework = "lambda"`, `criterion = "elasticity"`
3. `framework = "R0"`, `criterion = "standard"`
4. `framework = "R0"`, `criterion = "elasticity"`

In the \(\lambda\) framework, we consider both standard aggregation and elasticity-consistent 
aggregation. Standard aggregation follows the method of Hooley (2000), while elasticity-consistent 
aggregation uses the "genealogical collapsing" method of Bienvenu et al. (2017). Both approaches 
maintain the consistency of the asymptotic population growth rate \(\lambda\) and 
stable stage distribution. Elasticity-consistent aggregation additionally maintains
consistency of the reproductive values and the elasticities of \(\lambda\) with respect 
to matrix entries.

We extend these approaches to the \(R_0\) framework, where standard aggregation 
maintains the consistency of the net reproductive rate \(R_0\) and the cohort stable 
stage distribution. Elasticity-consistent aggregation in this framework additionally 
maintains consistency of the cohort reproductive values and the elasticities of \(R_0\) with respect 
to matrix entries. 

```{r aggregate}
agg1 <- mpm_aggregate(
  matU = matU,
  matF = matF,
  groups = groups,
  framework = "lambda",
  criterion = "standard"
)

agg2 <- mpm_aggregate(
  matU = matU,
  matF = matF,
  groups = groups,
  framework = "lambda",
  criterion = "elasticity"
)

agg3 <- mpm_aggregate(
  matU = matU,
  matF = matF,
  groups = groups,
  framework = "R0",
  criterion = "standard"
)

agg4 <- mpm_aggregate(
  matU = matU,
  matF = matF,
  groups = groups,
  framework = "R0",
  criterion = "elasticity"
)

# Extract aggregated U and F robustly (in case object field names differ by version)
extract_U <- function(x) x$matUk_agg %||% x$matU_agg %||% x$matUagg %||% x$U
extract_F <- function(x) x$matFk_agg %||% x$matF_agg %||% x$matFagg %||% x$F
extract_A <- function(x) x$matAk_agg %||% x$matA_agg %||% x$matAagg %||% x$A

as_ufA <- function(x) {
  U <- extract_U(x)
  F <- extract_F(x)
  A <- extract_A(x)
  if (is.null(A) && !is.null(U) && !is.null(F)) A <- U + F
  list(U = U, F = F, A = A)
}

m0 <- list(U = matU, F = matF, A = matA)
m1 <- as_ufA(agg1)
m2 <- as_ufA(agg2)
m3 <- as_ufA(agg3)
m4 <- as_ufA(agg4)


eff <- c(NA,agg1$effectiveness,agg2$effectiveness,agg3$effectiveness,agg4$effectiveness)

models <- list(
  "Original"                  = m0,
  "Agg lambda + standard"     = m1,
  "Agg lambda + elasticity"   = m2,
  "Agg R0 + standard"         = m3,
  "Agg R0 + elasticity"       = m4
)

model_labels <- c(
  Original         = "Original",
  Agg_lambda_std   = "Agg &lambda; + standard",
  Agg_lambda_elast = "Agg &lambda; + elasticity",
  Agg_R0_std       = "Agg R<sub>0</sub> + standard",
  Agg_R0_elast     = "Agg R<sub>0</sub> + elasticity"
)

names(models) <- c(
  "Original",
  "Agg_lambda_std",
  "Agg_lambda_elast",
  "Agg_R0_std",
  "Agg_R0_elast"
)

# Basic checks
stopifnot(all(sapply(models, function(m) !is.null(m$A))))
#stopifnot(all(sapply(models, function(m) nrow(m$A) == length(groups))))

#show aggregated matrices
print("Agg λ + standard")
m1$A
print("Agg λ + elasticity")
m2$A
print("Agg R0 + standard")
m3$A
print("Agg R0 + elasticity")
m4$A

```

**Important.** Elasticity-consistent aggregation can, in some cases, produce biologically 
infeasible parameter values (Bienvenu et al. 2017). As seen in the aggregated matrices above, 
aggregation under the \(\lambda\) framework yields a model with a survival probability 
greater than 1 (1.12). A similar issue arises under the \(R_0\) framework, where 
elasticity-consistent aggregation results in a survival probability of 1.22. This issue 
does not occur when using standard aggregation in the \(\lambda\) framework 
(Hooley, 2000) or in the \(R_0\) framework.

## Compare \(\lambda\), \(R_0\), \(T_a\), and \(T_c\) in a table

```{r results-table}

results <- data.frame(
  Model  = unname(model_labels[names(models)]),
  lambda = sapply(models, function(m) spectral_radius(m$A)),
  R0     = sapply(models, function(m) get_R0(m$U, m$F)),
  Ta     = sapply(models, function(m) get_Ta(m$U, m$F)),
  Tc     = sapply(models, function(m) get_Tc(m$U, m$F)),
  Effectiveness = eff,
  row.names = NULL,
  stringsAsFactors = FALSE
)

n_rows <- nrow(results)

knitr::kable(
  results,
  col.names = c(
    "Model",
    "<em>&lambda;</em>",
    "<em>R</em><sub>0</sub>",
    "<em>T</em><sub>a</sub> (years)",
    "<em>T</em><sub>c</sub> (years)",
    "<em>&rho;</em><sup>2</sup>"
  ),
  digits = 3,
  caption = paste0(
  "<strong>Table 1.</strong> Comparison of demographic properties for the original ",
  "stage-structured matrix population model of <em>Rourea induta</em> ",
  "(COMPADRE MatrixID 246781) and four aggregated versions. The original 14 stages ",
  "are collapsed to 5 while leaving the Seedling and Sucker stages unchanged. ",
  "&ldquo;Standard&rdquo; denotes use of a standard aggregator and ",
  "&ldquo;elasticity&rdquo; denotes elasticity-consistent aggregation. ",
  "<em>T</em><sub>a</sub> is generation time (years) in the <em>&lambda;</em> framework and ",
  "<em>T</em><sub>c</sub> is cohort generation time (years) in the ",
  "<em>R</em><sub>0</sub> framework. ",
  "<em>&rho;</em><sup>2</sup> quantifies aggregation effectiveness, with values closer to 1 ",
  "indicating closer agreement between the aggregated and reference models."
  ),
  format = "html",
  escape = FALSE
) |>
  kable_styling(full_width = TRUE) |>
  row_spec(0, extra_css = "border-bottom: 2px solid black;") |>
  row_spec(nrow(results), extra_css = "border-bottom: 2px solid black;") |>
  footnote(
    general_title = "",
    general = paste0(
      "<span style='font-size: 90%;'>",
      "<strong>Note.</strong> The projection interval associated with the population ",
      "growth rate (<em>&lambda;</em>) is 1 year.</span>"
    ),
    footnote_as_chunk = TRUE,
    escape = FALSE
  )
```


**How to read Table 1.** The first row reports the true demographic parameters of 
the original matrix population model for *Rourea induta* (Hoffmann 1999), while the 
remaining rows report estimates obtained from aggregated versions of the MPM. 
Generation time is preserved under “Agg λ + elasticity,” and cohort generation time 
is preserved under “Agg R₀ + elasticity.” All aggregated models achieve near-perfect 
aggregation, with effectiveness values exceeding 0.99.

### Interpretation notes

- Aggregation under the \(\lambda\) framework prioritizes preservation of long-term population growth.
- Aggregation under the \(R_0\) framework emphasizes lifetime reproductive output.

## Elasticities of \(\lambda\): making them comparable across models

To make the elasticities of the original model comparable with those of the
aggregated models, we transform the original elasticity matrix to the
aggregated stage structure as

\[
\mathbf{E}^{\mathrm{true}}_{\mathrm{agg}}
=
\mathbf{P}\, \mathbf{E}\!\bigl(\mathbf{A}\bigr)\, \mathbf{P}^{\top},
\]

where \( \mathbf{E}\!\bigl(\mathbf{A}\bigr) \) is the elasticity of
\( \lambda \) with respect to the entries of the original projection
matrix \( \mathbf{A} \), and \( \mathbf{P} \) is the partitioning matrix
induced by `groups`. Elasticities for each aggregated model are then
computed directly from its aggregated projection matrix, and the results
for all models are plotted together.

```{r elasticities}
P <- mpm_partition(groups=groups, n=nrow(matA))
# Elasticity matrices in aggregated space (k x k) for each model
E_A <- elasticity(matA=matA,framework="lambda")$elasticity
E_list <- list(
  "Original" = P %*% E_A %*% t(P),
  "Agg lambda + standard" = elasticity(matA=m1$A,framework="lambda")$elasticity,
  "Agg lambda + elasticity" = elasticity(matA=m2$A,framework="lambda")$elasticity,
  "Agg R0 + standard" = elasticity(matA=m3$A,framework="lambda")$elasticity,
  "Agg R0 + elasticity" = elasticity(matA=m4$A,framework="lambda")$elasticity
)

# Build a long data.frame for all nonzero entries (row-major order)
to_long <- function(M, model_name) {
  idx <- which(M != 0, arr.ind = TRUE)
  data.frame(
    model = model_name,
    row = idx[, 1],
    col = idx[, 2],
    entry = paste(idx[, 1], idx[, 2], sep = ","),
    elasticity = M[idx],
    stringsAsFactors = FALSE
  )
}

elast_df <- do.call(rbind, Map(to_long, E_list, names(E_list)))

# Keep positive values for log scale
elast_df <- elast_df[elast_df$elasticity > 0, ]

# Row-major ordering: row 1 entries first, row 2 second, rows 3+ third, etc.
elast_df <- elast_df[order(elast_df$row, elast_df$col), ]

models_order <- names(E_list)
entries <- unique(elast_df$entry)

# Create matrix for barplot: rows = models (5), cols = entries
elast_mat <- sapply(models_order, function(m) {
  elast_df$elasticity[elast_df$model == m]
})
elast_mat <- t(elast_mat)
rownames(elast_mat) <- models_order
colnames(elast_mat) <- entries
```

### Plot elasticities of \(\lambda\) by matrix entry

Each x-axis category corresponds to a matrix entry \([i,j]\) that is nonzero in at least 
one of the five projection matrices. Within each category, elasticities are shown for 
the original model and the four aggregated models.

**Figure 1. Elasticities of \(\lambda\) with respect to matrix entries.**  
Bars are grouped by matrix entry, with colors indicating the different five models. Because 
elasticities sum to 1 within each matrix, the figure illustrates how the influence on 
\(\lambda\) is distributed across life-history components. The models shown in the 
legend correspond to those described in Table 1.


```{r lambda-elasticity-plot, fig.width=12}
# Build biologically meaningful entry labels
# Build (row, col) vectors in the exact column order of elast_mat
rc <- do.call(rbind, strsplit(colnames(elast_mat), ","))
r_vec <- as.integer(rc[, 1])
c_vec <- as.integer(rc[, 2])

make_entry_label <- function(r, c) {
  if (r == 1) {
    bquote(F[1, .(c)])
  } else if (r == 2) {
    bquote(C[2, .(c)])
  } else {
    bquote(U[.(r), .(c)])
  }
}

entry_labels <- mapply(make_entry_label, r_vec, c_vec, SIMPLIFY = FALSE)


# Colors by model
# Same colors as the vital-rate plot for consistency
cols <- c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")

# Leave room for left legend and long x labels
op <- par(mar = c(7, 6, 4, 2))

bp <- barplot(
  elast_mat,
  beside = TRUE,
  log = "y",
  col = cols,
  border = NA,
  ylab = expression("Elasticity of " * lambda * " (log scale)"),
  xaxt = "n",
  space = c(0.2, 1)   # ← this is the key line
)

axis(
  side = 1,
  at = colMeans(bp),
  labels = entry_labels,
  las = 2,
  cex.axis = 0.9
)

legend(
  "topleft",
  legend = c(
    "Original",
    expression(paste("Agg ", lambda, " + standard")),
    expression(paste("Agg ", lambda, " + elasticity")),
    expression(paste("Agg ", R[0], " + standard")),
    expression(paste("Agg ", R[0], " + elasticity"))
  ),
  fill = cols,
  bty = "n",
  inset = 0.02
)

par(op)
```

**How to read Figure 1.** Each x-axis label corresponds to a vital rate represented by 
a matrix entry \([i,j]\). Fertility entries \(F[i,j]\) describe sexual reproduction, 
representing the contribution of individuals in stage \(i\) to seedling production 
over the projection interval. Clonal reproduction entries \(C[i,j]\) represent the 
contribution of individuals in stage \(i\) to the production of suckers over the 
projection interval. Survival and transition entries \(U[i,j]\) give the probability 
that an individual in stage \(i\) survives and transitions to stage \(j\) over the 
projection interval.


## Elasticities of \(R_0\): making them comparable across models

To make the elasticities of the original model comparable with those of the
aggregated models, we transform the original elasticity matrix to the
aggregated stage structure as

\[
\mathbf{E}^{\mathrm{true}}_{\mathrm{agg}}
=
\mathbf{P}\, \mathbf{E}\!\bigl(\mathbf{A}\bigr)\, \mathbf{P}^{\top},
\]

where \( \mathbf{E}\!\bigl(\mathbf{A}\bigr) \) is the elasticity of
\( R_0 \) with respect to the entries of the original projection
matrix \( \mathbf{A} \), and \( \mathbf{P} \) is the partitioning matrix
induced by `groups`. Elasticities for each aggregated model are then
computed directly from its aggregated projection matrix, and the results
for all models are plotted together.

```{r elasticities of R0}
P <- mpm_partition(groups=groups, n=nrow(matA))
# Elasticity matrices in aggregated space (k x k) for each model
E_A <- elasticity(matF=matF,matU=matU,framework="R0", normalize=TRUE)$elasticity
E_list <- list(
  "Original" = P %*% E_A %*% t(P),
  "Agg lambda + standard" = elasticity(matF=m1$F, matU=m1$U,framework="R0", normalize=TRUE)$elasticity,
  "Agg lambda + elasticity" = elasticity(matF=m2$F, matU=m2$U,framework="R0", normalize=TRUE)$elasticity,
  "Agg R0 + standard" = elasticity(matF=m3$F, matU=m3$U,framework="R0", normalize=TRUE)$elasticity,
  "Agg R0 + elasticity" = elasticity(matF=m4$F, matU=m4$U,framework="R0", normalize=TRUE)$elasticity
)

# Build a long data.frame for all nonzero entries (row-major order)
to_long <- function(M, model_name) {
  idx <- which(M != 0, arr.ind = TRUE)
  data.frame(
    model = model_name,
    row = idx[, 1],
    col = idx[, 2],
    entry = paste(idx[, 1], idx[, 2], sep = ","),
    elasticity = M[idx],
    stringsAsFactors = FALSE
  )
}

elast_df <- do.call(rbind, Map(to_long, E_list, names(E_list)))

# Keep positive values for log scale
elast_df <- elast_df[elast_df$elasticity > 0, ]

# Row-major ordering: row 1 entries first, row 2 second, rows 3+ third, etc.
elast_df <- elast_df[order(elast_df$row, elast_df$col), ]

models_order <- names(E_list)
entries <- unique(elast_df$entry)

# Create matrix for barplot: rows = models (5), cols = entries
elast_mat2 <- sapply(models_order, function(m) {
  elast_df$elasticity[elast_df$model == m]
})
elast_mat2 <- t(elast_mat2)
rownames(elast_mat2) <- models_order
colnames(elast_mat2) <- entries
```

### Plot elasticities of \(R_0\) by matrix entry

Each x-axis category corresponds to a matrix entry \([i,j]\) that is nonzero in at least 
one of the five projection matrices. Within each category, elasticities are shown for 
the original model and the four aggregated models.

**Figure 2. Elasticities of \(R_0\) with respect to matrix entries.**  
Bars are grouped by matrix entry, with colors indicating the five different models. Elasticities 
are normalized to sum to 1 within each matrix, so the figure illustrates how the 
contribution to \(R_0\) is distributed across life-history components. The models shown 
in the legend correspond to those described in Table 1.


```{r elasticity-R0-plot, fig.width=12, fig.height=6}
# Build biologically meaningful entry labels
# Build (row, col) vectors in the exact column order of elast_mat
rc <- do.call(rbind, strsplit(colnames(elast_mat2), ","))
r_vec <- as.integer(rc[, 1])
c_vec <- as.integer(rc[, 2])

make_entry_label <- function(r, c) {
  if (r == 1) {
    bquote(F[1, .(c)])
  } else if (r == 2) {
    bquote(C[2, .(c)])
  } else {
    bquote(U[.(r), .(c)])
  }
}

entry_labels <- mapply(make_entry_label, r_vec, c_vec, SIMPLIFY = FALSE)


# Colors by model
# Same colors as the vital-rate plot for consistency
cols <- c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")

# Leave room for left legend and long x labels
op <- par(mar = c(7, 6, 4, 2))

bp <- barplot(
  elast_mat2,
  beside = TRUE,
  log = "y",
  col = cols,
  border = NA,
  ylab = expression("Normalized elasticity of " * R[0] * " (log scale)"),
  xaxt = "n",
  space = c(0.2, 1)   # ← this is the key line
)

axis(
  side = 1,
  at = colMeans(bp),
  labels = entry_labels,
  las = 2,
  cex.axis = 0.9
)

legend(
  "topleft",
  legend = c(
    "Original",
    expression(paste("Agg ", lambda, " + standard")),
    expression(paste("Agg ", lambda, " + elasticity")),
    expression(paste("Agg ", R[0], " + standard")),
    expression(paste("Agg ", R[0], " + elasticity"))
  ),
  fill = cols,
  bty = "n",
  inset = 0.02
)

par(op)
```
**How to read Figure 2.** Each x-axis label corresponds to a vital rate represented by 
a matrix entry \([i,j]\). Fertility entries \(F[i,j]\) describe sexual reproduction, 
representing the contribution of individuals in stage \(i\) to seedling production 
over the projection interval. Clonal reproduction entries \(C[i,j]\) represent the 
contribution of individuals in stage \(i\) to the production of suckers over the 
projection interval. Survival and transition entries \(U[i,j]\) give the probability 
that an individual in stage \(i\) survives and transitions to stage \(j\) over the 
projection interval.

Across both Figures 1 and 2, the majority of elasticity is concentrated in a small 
number of vital rates. In particular, three survival terms—\(U[3,3]\), \(U[4,4]\), 
and \(U[5,5]\)—account for most of the total elasticity. These entries 
correspond to stasis within the larger size classes and indicate that population 
performance, whether measured by \(\lambda\) or \(R_0\), is driven primarily by 
persistence in these stages. This pattern is consistent across the original and 
aggregated models, suggesting that aggregation preserves the dominant contributors 
to asymptotic population dynamics.

## Comparing elasticities of \(\lambda\) and \(R_0\)
There is remarkable consistency between the elasticities in the \(\lambda\) and \(R_0\) 
frameworks. In both cases, the same three matrix entries account for the majority of elasticity 
mass, indicating strong agreement in the demographic processes identified as most 
influential. These shared entries account for 96% of the total elasticity mass in the 
\(\lambda\) framework and 94% in the \(R_0\) framework. Elasticities in the \(R_0\) 
framework are normalized to sum to 1, whereas elasticities in the \(\lambda\) framework 
require no such normalization because they already sum to 1 by definition.

```{r lambda-vs.-R0-elasticities}
#compare elasticities of top 3 vital rates
#top three elasticities in lambda framework
sum(elast_mat[,c(9,13,15)])/5
#top three elasticities in R0 framework
sum(elast_mat2[,c(9,13,15)])/5
```
## Summary

Across aggregation methods, estimates of \(\lambda\) and \(R_0\) in Table 1 are broadly 
similar, indicating that aggregation preserves overall conclusions about population 
growth. Differences among methods are small and largely reflect whether aggregation 
prioritizes long-term growth (\(\lambda\)) or lifetime reproduction (\(R_0\)). In 
contrast, generation time estimates clearly differ between frameworks: generation 
time in the \(\lambda\) framework (\(T_a\)) and cohort generation time in the \(R_0\) 
framework (\(T_c\)) reflect fundamentally different perspectives on the pace of life.

Particularly striking is that \(\lambda\) is close to 1 (within 1%), yet 
\(R_0\) is not. If \(\lambda\) were exactly equal to 1, then \(R_0\) would also necessarily 
equal 1, and all corresponding demographic quantities would coincide across the two frameworks. 
That the demographic quantities differ greatly here, reveals how sensitive demographic 
parameters can sometimes be to even small deviations of the stable growth rate around 
unity.

Despite these differences in scalar demographic quantities, elasticity patterns are 
remarkably consistent across frameworks. Elasticities of \(\lambda\) and 
\(R_0\) identify the same matrix entries as dominant contributors, with most of the elasticity 
mass shared between them. This consistency suggests that, in this example, the 
demographic processes driving population growth and reproductive output are essentially 
the same, even when different aggregation objectives are used.

## References

Bienvenu, F., Akçay, E., Legendre, S. and McCandlish, D.M. 2017. The genealogical
decomposition of a matrix population model with applications to the aggregation
of stages. Theoretical Population Biology, 115, 69-80. 
https://doi.org/10.1016/j.tpb.2017.04.002

Hoffmann, W.A., 1999. Fire and population dynamics of woody plants in a neotropical 
savanna: matrix model projections. Ecology, 80(4), 1354-1369.
https://doi.org/10.1890/0012-9658(1999)080[1354:FAPDOW]2.0.CO;2

Hooley, D. E. 2000. Collapsed matrices with (almost) the same eigenstuff. The 
College Mathematics Journal, 31(4), 297–299. https://doi.org/10.1080/07468342.2000.11974162

Jones, O. R., Barks, P., Stott, I., James, T. D., Levin, S., Petry, W. K., Capdevila, P., 
Che-Castaldo, J., Jackson, J., Römer, G., Schuette, C., Thomas, C. C., & Salguero-Gómez, R. 2022. 
Rcompadre and Rage—Two R packages to facilitate the use of the COMPADRE and COMADRE
databases and calculation of life-history traits from matrix population models. 
Methods in Ecology and Evolution, 13(4), 770–781. https://doi.org/10.1111/2041-210X.13792

Salguero-Gómez R., & Plotkin, J. B. (2010). Matrix dimensions bias demographic 
inferences: implications for comparative plant demography. The American Naturalist, 
176(6), 710–722. https://doi.org/10.1086/657044

Salguero-Gómez, R., Jones, O. R., Archer, C. R., Buckley, Y. M., Che-Castaldo, J., 
Caswell, H., Hodgson, D., Scheuerlein, A., Conde, D. A., Brinks, E., de Buhr, H., 
Farack, C., Gottschalk, F., Hartmann, A., Henning, A., Hoppe, G., Römer, G., 
Runge, J., Ruoff, T., Wille, J., Zeh, S., Davison, R., Vieregg, D., Baudisch, A., 
Altwegg, R., Colchero, F., Dong, M., de Kroon, H., Lebreton, J. -D., Metcalf, C. J. E., 
Neel, M. M., Parker, I. M., Takada, T., Valverde, T., Vélez-Espino, L. A., Wardle, G. M., 
Franco, M., & Vaupel, J. W. (2015). The COMPADRE Plant Matrix Database: an open 
online repository for plant demography. Journal of Ecology, 103, 202–218. 
https://doi.org/10.1111/1365-2745.12334
