---
title: "Aggregating a Leslie-to-Leslie MPM"
author:
  - name: "Richard A. Hinrichsen"
    orcid: "0000-0003-0761-3005"
output:
  rmarkdown::html_vignette:
    number_sections: false
    mathjax: default
vignette: >
  %\VignetteIndexEntry{Aggregating a Leslie-to-Leslie MPM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**ORCID:** <https://orcid.org/0000-0003-0761-3005>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

Matrix population models (MPMs) are often constructed at different levels of age 
or stage resolution, which can complicate comparison across models. For age-structured 
populations, this commonly occurs when a finely resolved Leslie matrix must be compared 
with a coarser age-structured representation. In this setting, aggregation can be 
used to collapse fine age classes into a smaller number of broader age groups, 
yielding a coarser Leslie matrix that is directly comparable to the original while 
retaining its key demographic properties.

In this vignette, we demonstrate how to aggregate a Leslie matrix population model to
another Leslie matrix population model using the `mpmaggregate` package. Leslie-to-Leslie
aggregation reconciles differences in age resolution while keeping the model within the
Leslie class of age-structured projection matrices.

As a worked example, we use an age-structured matrix population model for the
yellow-bellied marmot *Marmota flaviventris* obtained from the COMADRE database
(Jones et al. 2022), which archives demographic data originally published by
Ozgul et al. (2009). We focus on aggregation using `leslie_aggregate()`, and we compare
four combinations of demographic framework and aggregation criterion, highlighting what
each approach preserves and how the aggregated models differ in interpretation.

## Packages

We begin by loading the packages used to access matrix population models, perform matrix
aggregation, and format output for display.

```{r packages}
library(mpmaggregate)
library(kableExtra)
library(rphylopic)
library(expm)
```

## Data acquisition

To demonstrate Leslie-to-Leslie aggregation, we use a 10 × 10 age-structured matrix
population model for *Marmota flaviventris* (MatrixID **249455**) from the COMADRE database
(Ozgul et al. 2009). The model contains multiple fine-grained age classes, making it a
suitable candidate for Leslie-to-Leslie aggregation (Hinrichsen, 2023; Hinrichsen et al., 2026).

The yellow-bellied marmot *Marmota flaviventris* is a large, stout-bodied ground squirrel
(rodent) native to the mountainous regions of western North America.

```{r}
# Leslie population projection matrix for Marmota flaviventris
# MatrixID = 249455

# Matrices can be retrieved with:
# library(Rcompadre)
# comadre <- cdb_fetch("comadre",version = "4.23.3.1")
# mpm <- comadre[comadre$MatrixID == 249455, ]
# matA <- matA(mpm)[[1]]
# matU <- matU(mpm)[[1]]
# matF <- matF(mpm)[[1]]

# Not run: requires Rcompadre and internet access

# Leslie population projection matrix
matA <- matrix(c(
  0.000000, 0.258040, 0.627254, 0.700037, 0.700037, 0.700037, 0.700037, 0.700037, 0.700037, 0.700037,
  0.545125, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000,
  0.000000, 0.498625, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000,
  0.000000, 0.000000, 0.592875, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000,
  0.000000, 0.000000, 0.000000, 0.643500, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000,
  0.000000, 0.000000, 0.000000, 0.000000, 0.643500, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000,
  0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.643500, 0.000000, 0.000000, 0.000000, 0.000000,
  0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.643500, 0.000000, 0.000000, 0.000000,
  0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.643500, 0.000000, 0.000000,
  0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.000000, 0.643500, 0.000000
), nrow = 10, byrow = TRUE)


# Reproduction matrix
matF <- matrix(c(
  0, 0.25804, 0.627254, 0.700037, 0.700037, 0.700037, 0.700037, 0.700037, 0.700037, 0.700037,
  0, 0,       0,        0,        0,        0,        0,        0,        0,        0,
  0, 0,       0,        0,        0,        0,        0,        0,        0,        0,
  0, 0,       0,        0,        0,        0,        0,        0,        0,        0,
  0, 0,       0,        0,        0,        0,        0,        0,        0,        0,
  0, 0,       0,        0,        0,        0,        0,        0,        0,        0,
  0, 0,       0,        0,        0,        0,        0,        0,        0,        0,
  0, 0,       0,        0,        0,        0,        0,        0,        0,        0,
  0, 0,       0,        0,        0,        0,        0,        0,        0,        0,
  0, 0,       0,        0,        0,        0,        0,        0,        0,        0
), nrow = 10, byrow = TRUE)

# Survival/transition matrix
matU <- matrix(c(
  0,       0,       0,       0,      0,      0,      0,      0,      0,      0,
  0.545125,0,       0,       0,      0,      0,      0,      0,      0,      0,
  0,       0.498625,0,       0,      0,      0,      0,      0,      0,      0,
  0,       0,       0.592875,0,      0,      0,      0,      0,      0,      0,
  0,       0,       0,       0.6435, 0,      0,      0,      0,      0,      0,
  0,       0,       0,       0,      0.6435, 0,      0,      0,      0,      0,
  0,       0,       0,       0,      0,      0.6435, 0,      0,      0,      0,
  0,       0,       0,       0,      0,      0,      0.6435, 0,      0,      0,
  0,       0,       0,       0,      0,      0,      0,      0.6435, 0,      0,
  0,       0,       0,       0,      0,      0,      0,      0,      0.6435, 0
), nrow = 10, byrow = TRUE)

# Sanity check: A = U + F (after folding any C into F)
stopifnot(all.equal(unname(matA), unname(matU + matF)))
   
```

## Helper functions for demographic quantities

We compute:

- \( \lambda \): spectral radius of \( \mathbf{A} \)
- \( R_0 \): spectral radius of \( \mathbf{F} \left( \mathbf{I} - \mathbf{U} \right)^{-1} \)
- \( T_a \): generation time (years) under the \( \lambda \) framework
- \( T_c \): cohort generation time (years) under the \( R_0 \) framework

```{r helpers}
`%||%` <- function(x, y) if (!is.null(x)) x else y

get_R0 <- function(U, F) {
  I <- diag(nrow(U))
  N <- solve(I - U)
  K <- F %*% N
  spectral_radius(K)
}

get_Ta <- function(U, F) {
  generation_time(F, U, framework = "lambda")$generation_time
}

get_Tc <- function(U, F) {
  generation_time(F, U, framework = "R0")$generation_time
}
```

## Aggregate in four ways

We aggregate the model using the following four combinations of framework and criterion:

1. `framework = "lambda"`, `criterion = "standard"`
2. `framework = "lambda"`, `criterion = "elasticity"`
3. `framework = "R0"`, `criterion = "standard"`
4. `framework = "R0"`, `criterion = "elasticity"`

In the \(\lambda\) framework, standard aggregation follows the method of Hooley (2000),
while elasticity-consistent aggregation was first described by Bienvenu et al. (2017)
and extended to handle Leslie-to-Leslie aggregation (Hinrichsen, 2023; Hinrichsen et al., 2026).
Both approaches maintain  consistency of the asymptotic growth rate \(\lambda\) and the stable
age distribution. Elasticity-consistent aggregation additionally maintains consistency
of the reproductive values and the elasticities of \(\lambda\) with respect to entries of 
\( \mathbf{A}^k \), where \(k = n/m\).

We extend these approaches to the \(R_0\) framework, where standard aggregation maintains the 
consistency of the net reproductive rate \(R_0\) and cohort stable age distribution.
Elasticity-consistent aggregation in this framework additionally maintains the consistency of
the cohort reproductive values and elasticities of \(R_0\) with respect to matrix entries.

```{r aggregate}
# Aggregate into 5 x 5 matrices (m = 5)
ngroups <- 5

agg1 <- leslie_aggregate(
  matA = matA, ngroups = ngroups,
  framework = "lambda", criterion = "standard"
)

agg2 <- leslie_aggregate(
  matA = matA, ngroups = ngroups,
  framework = "lambda", criterion = "elasticity"
)

agg3 <- leslie_aggregate(
  matA = matA, ngroups = ngroups,
  framework = "R0", criterion = "standard"
)

agg4 <- leslie_aggregate(
  matA = matA, ngroups = ngroups,
  framework = "R0", criterion = "elasticity"
)

# Split a Leslie matrix into U and F components (Leslie form: fertility in row 1)
split_matA <- function(A) {
  n <- nrow(A)
  U <- A
  U[1, ] <- 0
  F <- matrix(0, n, n)
  F[1, ] <- A[1, ]
  list(U = U, F = F, A = A)
}

m0 <- list(U = matU, F = matF, A = matA)
m1 <- split_matA(agg1$matAk_agg)
m2 <- split_matA(agg2$matAk_agg)
m3 <- split_matA(agg3$matAk_agg)
m4 <- split_matA(agg4$matAk_agg)

eff <- c(
  NA,
  agg1$effectiveness,
  agg2$effectiveness,
  agg3$effectiveness,
  agg4$effectiveness
)

models <- list(
  Original            = m0,
  Agg_lambda_std      = m1,
  Agg_lambda_elast    = m2,
  Agg_R0_std          = m3,
  Agg_R0_elast        = m4
)

model_labels <- c(
  Original         = "Original",
  Agg_lambda_std   = "Agg &lambda; + standard",
  Agg_lambda_elast = "Agg &lambda; + elasticity",
  Agg_R0_std       = "Agg R<sub>0</sub> + standard",
  Agg_R0_elast     = "Agg R<sub>0</sub> + elasticity"
)

# Basic checks
stopifnot(all(sapply(models, function(m) !is.null(m$A))))

# Show aggregated matrices
print("Agg λ + standard");   m1$A
print("Agg λ + elasticity"); m2$A
print("Agg R0 + standard");  m3$A
print("Agg R0 + elasticity");m4$A
```

**Important.** Unlike general-to-general aggregation, Leslie-to-Leslie aggregation cannot
produce survival probabilities greater than 1. It is always true that the survival
probabilities are the same under standard and elasticity-consistent aggregation within each
framework, while the fertility schedules can differ between the standard and elasticity-consistent
aggregated Leslie matrices (Hinrichsen et al., 2026).

## Compare \(\lambda\), \(R_0\), \(T_a\), and \(T_c\) in a table

```{r results-table alt}
# Adjust lambdas so that they are all per capita growth rate over a year
# Adjust generation times so they are all based on years
adj = c(1,2,2,2,2)

lambda = sapply(models, function(m) spectral_radius(m$A))
lambda = lambda^(1/adj)
R0     = sapply(models, function(m) get_R0(m$U, m$F))
Ta     = sapply(models, function(m) get_Ta(m$U, m$F))*adj
Tc     = sapply(models, function(m) get_Tc(m$U, m$F))*adj

results <- data.frame(
  Model  = unname(model_labels[names(models)]),
  lambda = lambda,
  R0     = R0,
  Ta     = Ta,
  Tc     = Tc,
  Effectiveness = eff,
  row.names = NULL,
  stringsAsFactors = FALSE
)

n_rows <- nrow(results)

knitr::kable(
  results,
  col.names = c(
    "Model",
    "<em>&lambda;</em>",
    "<em>R</em><sub>0</sub>",
    "<em>T</em><sub>a</sub> (years)",
    "<em>T</em><sub>c</sub> (years)",
    "<em>&rho;</em><sup>2</sup>"
  ),
  digits = 3,
  caption = paste0(
    "<strong>Table 1.</strong> Comparison of demographic properties for the ",
    "yellow-bellied marmot <em>Marmota flaviventris</em> (COMADRE MatrixID 249455) ",
    "using the original 10 × 10 Leslie matrix and four aggregated Leslie matrices ",
    "collapsed to 5 age groups. Metrics include population growth rate ",
    "(<em>&lambda;</em>), net reproductive rate (<em>R</em><sub>0</sub>), ",
    "generation time (years; <em>T</em><sub>a</sub>), cohort generation time ",
    "(years; <em>T</em><sub>c</sub>), and aggregation effectiveness ",
    "(<em>&rho;</em><sup>2</sup>). &ldquo;Standard&rdquo; denotes use of a standard aggregator ",
    "and &ldquo;elasticity&rdquo; denotes elasticity-consistent aggregation."
  ),
  format = "html",
  escape = FALSE
) |>
  kable_styling(full_width = TRUE) |>
  row_spec(0, extra_css = "border-bottom: 2px solid black;") |>
  row_spec(nrow(results), extra_css = "border-bottom: 2px solid black;") |>
  footnote(
    general_title = "",
    general = paste0(
      "<span style='font-size: 90%;'>",
      "<strong>Note.</strong> Population growth rates (<em>&lambda;</em>) are all transformed to ",
      "represent growth over 1 year. Generation times have all been transformed to ",
      "represent generation time measured in years.</span>"
    ),
    footnote_as_chunk = TRUE,
    escape = FALSE
  )
```

**How to read Table 1.** The first row reports demographic quantities computed from the
original Leslie matrix for *Marmota flaviventris* (Ozgul et al. 2009), while the remaining
rows report estimates obtained from aggregated Leslie matrices.

### Interpretation notes

- Aggregation under the **\(\lambda\) framework** prioritizes consistency of long-term population growth.
- Aggregation under the **\(R_0\) framework** prioritizes consistency of net reproductive rate.


## Elasticities of \(\lambda\): making them comparable across models

To make the elasticities of the original model comparable with those of the
aggregated models, we transform the original elasticity matrix to the
aggregated age structure as

\[
\mathbf{E}^{\mathrm{true}}_{\mathrm{agg}}
=
\mathbf{P}\, \mathbf{E}\!\bigl(\mathbf{A}^{k}\bigr)\, \mathbf{P}^{\top},
\]

where \( \mathbf{E}\!\bigl(\mathbf{A}^{k}\bigr) \) is the elasticity of
\( \lambda^{k} \) with respect to the entries of the original
\( \mathbf{A}^{k} \), and \( \mathbf{P} \) is the partitioning matrix
induced by `groups`. Elasticities for each aggregated model are then
computed directly from its aggregated projection matrix, and the results
for all models are plotted together.

```{r elasticities-lambda^k}

# Age groups used for aggregation in example
groups <- list(c(1,2), c(3,4), c(5,6), c(7,8), c(9,10))

# Partitioning matrix induced by the age groups.
# This matrix is constructed internally by leslie_aggregate(),
# but we construct it explicitly here so it can be used to
# project quantities from the original age-class space to the
# aggregated space.
P <- mpm_partition(groups = groups, n = nrow(matA))

# Dimensions of the original and aggregated matrices
m <- ngroups
n <- nrow(matA)
k <- n / m

matAk <- matA %^% k

# Elasticity of lambda^k with respect to entries of A^k (closest to what aggregated matrices yield)
E_A <- mpm_elasticity(matA = matAk, framework = "lambda")$elasticity

E_list <- list(
  "Original"            = P %*% E_A %*% t(P),
  "Agg λ + standard"    = mpm_elasticity(matA = m1$A, framework = "lambda")$elasticity,
  "Agg λ + elasticity"  = mpm_elasticity(matA = m2$A, framework = "lambda")$elasticity,
  "Agg R0 + standard"   = mpm_elasticity(matA = m3$A, framework = "lambda")$elasticity,
  "Agg R0 + elasticity" = mpm_elasticity(matA = m4$A, framework = "lambda")$elasticity
)

to_long <- function(M, model_name) {
  idx <- which(M != 0, arr.ind = TRUE)
  data.frame(
    model = model_name,
    row = idx[, 1],
    col = idx[, 2],
    entry = paste(idx[, 1], idx[, 2], sep = ","),
    elasticity = M[idx],
    stringsAsFactors = FALSE
  )
}

elast_df <- do.call(rbind, Map(to_long, E_list, names(E_list)))
elast_df <- elast_df[elast_df$elasticity > 0, ]
elast_df <- elast_df[order(elast_df$row, elast_df$col), ]

models_order <- names(E_list)
entries <- unique(elast_df$entry)

elast_mat <- sapply(models_order, function(m) elast_df$elasticity[elast_df$model == m])
elast_mat <- t(elast_mat)
rownames(elast_mat) <- models_order
colnames(elast_mat) <- entries
```

### Plot: elasticities of \( \lambda^k \) by matrix entry

Each x-axis category corresponds to a matrix entry \([i,j]\) that is nonzero in at least one of
the five \(A\) matrices. Within each category, we plot elasticities for the original
(“true aggregated”) model and the four aggregated models.

**Figure 1. Elasticities of \( \lambda^k \) with respect to matrix entries.**  
Elasticities quantify the proportional change in \( \lambda^k \) resulting from proportional
changes in individual vital rates. Bars are grouped by matrix entry, with colors indicating the
five models. Because elasticities sum to 1 within each matrix, the figure illustrates how the
influence on \( \lambda^k \) is distributed across life-history components. The models shown in
the legend correspond to those described in Table 1.

```{r elasticity-plot, fig.width=12, fig.height=6}
# Marmot 2012 Feb 2 by T. Michael Keesey, CC0 1.0, Marmota monax (Linnaeus 1758)
# uuid <- "eee50efb-40dc-47d0-b2cb-52a14a5e0e51"
# img <- get_phylopic(uuid = uuid, format = "raster", height = 869)
# Write image 
# png::writePNG(unclass(img), "inst/extdata/marmot.png")
# Not run: requires internet access

img_file <- system.file("extdata", "marmot.png", package = "mpmaggregate")

if (!nzchar(img_file)) {
  img_file <- "../inst/extdata/marmot.png"
}

img <- png::readPNG(img_file)

rc <- do.call(rbind, strsplit(colnames(elast_mat), ","))
r_vec <- as.integer(rc[, 1])
c_vec <- as.integer(rc[, 2])

make_entry_label <- function(r, c) {
  if (r == 1) {
    bquote(F[1, .(c)])
  } else {
    bquote(U[.(r), .(c)])
  }
}
entry_labels <- mapply(make_entry_label, r_vec, c_vec, SIMPLIFY = FALSE)

cols <- c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")
op <- par(mar = c(7, 6, 4, 2))

bp <- barplot(
  elast_mat,
  beside = TRUE,
  ylim = c(0, .35),
  col = cols,
  border = NA,
  ylab = expression("Elasticity of " * lambda^k),
  xaxt = "n",
  space = c(0.2, 1)
)

axis(side = 1, at = colMeans(bp), labels = entry_labels, las = 2, cex.axis = 0.9)
legend("topleft", legend = rownames(elast_mat), fill = cols, bty = "n", inset = 0.02)

lims <- par("usr")
mydiff <- lims[4] - lims[3]
add_phylopic_base(
  img = img,
  x = NULL,
  y = mydiff * .9,
  height = .2 * mydiff * dim(img)[1] / 1024
)

par(op)
```

**How to read Figure 1.** Each x-axis label corresponds to a vital rate represented by a matrix
entry \([i,j]\). For fertility, \(F[i,j]\) denotes reproduction over the projection interval. For
survival, \(U[i,j]\) gives the probability of surviving and transitioning from age class \(i\) to
age class \(j\) over the projection interval. Note that there is exact agreement between the
elasticities of models “Original” and “Agg \(\lambda\) + elasticity,” which is by design.
The silhouette is from PhyloPic (https://www.phylopic.org; Keesey, 2023) and were added 
using the rphylopic R package (Gearty & Jones, 2023). The silhouette was contributed as follows: 
*Marmota monax* by  T. Michael Keesey (2012; CC0 1.0).

## Elasticities of \(R_0\): making them comparable across models

To make the elasticities of the original model comparable with those of the
aggregated models, we transform the original elasticity matrix to the
aggregated stage structure as

\[
\mathbf{E}^{\mathrm{true}}_{\mathrm{agg}}
=
\mathbf{P}\, \mathbf{E}\!\left(\mathbf{A}_{k}\right)\, \mathbf{P}^{\top},
\]

where \( \mathbf{E}\!\left(\mathbf{A}_{k}\right) \) is the normalized elasticity of
\( R_0 \) with respect to the entries of the \( k \)-step-ahead projection matrix
\( \mathbf{A}_{k} \) in the \( R_0 \) framework, and \( \mathbf{P} \) is the partitioning
matrix induced by `groups`. Elasticities for each aggregated model are then
computed directly from its corresponding \( \mathbf{U}_{k} \) and
\( \mathbf{F}_{k} \), and the results for all models are plotted together.

```{r elasticities-r0}
# Age groupings used for aggregation (reintroduced here so this section
# can be read independently)
groups <- list(c(1,2), c(3,4), c(5,6), c(7,8), c(9,10))

# Partitioning matrix induced by the stage groups.
# Constructed internally by leslie_aggregate(), but reproduced here
# because it is used to map quantities to aggregated space.
P <- mpm_partition(groups = groups, n = nrow(matA))

# Dimensions of the original and aggregated matrices
# Reintroduced here for clarity
m <- ngroups
n <- nrow(matA)
k <- n / m

R0 <- spectral_radius(matF %*% solve(diag(n) - matU))

# Reference operator for aggregation eigenstructure in the R0 framework
# Divide by R0 so that it has Leslie matrix form
A_R0ref <- (matF + R0 * matU) / R0

Uk <- expm::`%^%`(matU, k)
Rk <- (expm::`%^%`(A_R0ref, k) - Uk) * R0
Ak <- Uk + Rk

# Elasticity of R0 with respect to entries of Ak (closest to what aggregated matrices yield)
E_A <- mpm_elasticity(matF = Rk, matU = Uk, framework = "R0", normalize = TRUE)$elasticity

E_list2 <- list(
  "Original"            = P %*% E_A %*% t(P),
  "Agg λ + standard"    = mpm_elasticity(matF = m1$F, matU = m1$U, framework = "R0", normalize = TRUE)$elasticity,
  "Agg λ + elasticity"  = mpm_elasticity(matF = m2$F, matU = m2$U, framework = "R0", normalize = TRUE)$elasticity,
  "Agg R0 + standard"   = mpm_elasticity(matF = m3$F, matU = m3$U, framework = "R0", normalize = TRUE)$elasticity,
  "Agg R0 + elasticity" = mpm_elasticity(matF = m4$F, matU = m4$U, framework = "R0", normalize = TRUE)$elasticity
)

elast_df2 <- do.call(rbind, Map(to_long, E_list2, names(E_list2)))
elast_df2 <- elast_df2[elast_df2$elasticity > 0, ]
elast_df2 <- elast_df2[order(elast_df2$row, elast_df2$col), ]

models_order2 <- names(E_list2)
entries2 <- unique(elast_df2$entry)

elast_mat2 <- sapply(models_order2, function(m) elast_df2$elasticity[elast_df2$model == m])
elast_mat2 <- t(elast_mat2)
rownames(elast_mat2) <- models_order2
colnames(elast_mat2) <- entries2
```

### Plot: elasticities of \(R_0\) by matrix entry

Each x-axis category corresponds to a matrix entry \([i,j]\) that is nonzero in at least one of
the five matrices. Within each category, we plot elasticities for the original
(“true aggregated”) model and the four aggregated models.

**Figure 2. Elasticities of \(R_0\) with respect to matrix entries.**  
Elasticities quantify the proportional change in \(R_0\) resulting from proportional changes in
individual vital rates. Bars are grouped by matrix entry, with colors indicating the five models.
The elasticities of \(R_0\) are normalized to sum to 1, so each elasticity can be interpreted as
the fraction of the total contribution to \(R_0\) attributable to that matrix entry.

```{r elasticity-r0-plot, fig.width=12, fig.height=6}
# Marmot 2012 Feb 2 by T. Michael Keesey, CC0 1.0, Marmota monax (Linnaeus 1758)
# uuid <- "eee50efb-40dc-47d0-b2cb-52a14a5e0e51"
# img <- get_phylopic(uuid = uuid, format = "raster", height = 869)
# Write image 
# png::writePNG(unclass(img), "inst/extdata/marmot.png")
# Not run: requires internet access

img_file <- system.file("extdata", "marmot.png", package = "mpmaggregate")

#For development
if (!nzchar(img_file)) {
  img_file <- "../inst/extdata/marmot.png"
}

img <- png::readPNG(img_file)

rc <- do.call(rbind, strsplit(colnames(elast_mat2), ","))
r_vec <- as.integer(rc[, 1])
c_vec <- as.integer(rc[, 2])

entry_labels <- mapply(make_entry_label, r_vec, c_vec, SIMPLIFY = FALSE)

cols <- c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")
op <- par(mar = c(7, 6, 4, 2))

bp <- barplot(
  elast_mat2,
  beside = TRUE,
  ylim = c(0, .35),
  col = cols,
  border = NA,
  ylab = expression("Normalized elasticity of " * R[0]),
  xaxt = "n",
  space = c(0.2, 1)
)

axis(side = 1, at = colMeans(bp), labels = entry_labels, las = 2, cex.axis = 0.9)
legend("topleft", legend = rownames(elast_mat2), fill = cols, bty = "n", inset = 0.02)

lims <- par("usr")
mydiff <- lims[4] - lims[3]
add_phylopic_base(
  img = img,
  x = NULL,
  y = mydiff * .9,
  height = .2 * mydiff * dim(img)[1] / 1024
)

par(op)
```

**How to read Figure 2.** Each x-axis label corresponds to a vital rate represented by a matrix
entry \([i,j]\). For fertility, \(F[i,j]\) denotes reproduction over the projection interval. For
survival, \(U[i,j]\) gives the probability of surviving and transitioning from age class \(i\) to
age class \(j\) over the projection interval. Note that there is exact agreement between the
elasticities of models “Original” and “Agg \(R_0\) + elasticity,” which is by design.
The silhouette is from PhyloPic (https://www.phylopic.org; Keesey, 2023) and were added 
using the rphylopic R package (Gearty & Jones, 2023). The silhouette was contributed as follows: 
*Marmota monax* by  T. Michael Keesey (2012; CC0 1.0).

### Comparing elasticities of \(\lambda^k\) and \(R_0\)

Within each framework, true elasticities and those derived from aggregated matrices show strong
qualitative agreement, regardless of aggregation method. Both approaches identify early survival
and early fertility as dominant contributors to population performance for both \(\lambda^k\) and
\(R_0\).

Overall, fertility plays a larger role in determining \(R_0\) than \(\lambda^k\), accounting for
approximately 51\% versus 45\% of total elasticity mass, respectively. Across analyses,
early survival probability (the transition \(U[2,1]\)) is the single most influential vital rate.

```{r fertility-elasticity-mass}
# Mass for elasticity of R0 w.r.t. fertilities
sum(elast_mat2[, 1:5]) / 5

# Mass for elasticity of lambda^k w.r.t. fertilities
sum(elast_mat[, 1:5]) / 5
```

## References

Bienvenu, F., Akçay, E., Legendre, S., and McCandlish, D. M. (2017). The genealogical
decomposition of a matrix population model with applications to the aggregation
of stages. *Theoretical Population Biology*, 115, 69–80.
https://doi.org/10.1016/j.tpb.2017.04.002

Gearty, W., and Jones, L. A. (2023). rphylopic: An R package for fetching,
transforming, and visualising PhyloPic silhouettes. 
*Methods in Ecology and Evolution*, 14, 2700–2708.
https://doi.org/10.1111/2041-210X.14221

Hooley, D. E. (2000). Collapsed matrices with (almost) the same eigenstuff.
*The College Mathematics Journal*, 31(4), 297–299.
https://doi.org/10.1080/07468342.2000.11974162

Hinrichsen, R. A. (2023). Aggregation of Leslie matrix models with application
to ten diverse animal species. *Population Ecology*, 65(3), 146–166.
https://doi.org/10.1002/1438-390X.12149

Hinrichsen, R. A., Yokomizo, H., and Salguero-Gómez, R. (2026). From theory to
application: Elasticity-consistent aggregation of Leslie matrix population
models for comparative demography. *bioRxiv*, preprint.
https://doi.org/10.64898/2026.02.04.703802

Jones, O. R., Barks, P., Stott, I., James, T. D., Levin, S., Petry, W. K.,
Capdevila, P., Che-Castaldo, J., Jackson, J., Römer, G., Schuette, C.,
Thomas, C. C., and Salguero-Gómez, R. (2022). Rcompadre and Rage—Two R
packages to facilitate the use of the COMPADRE and COMADRE databases and
calculation of life-history traits from matrix population models.
*Methods in Ecology and Evolution*, 13(4), 770–781.
https://doi.org/10.1111/2041-210X.13792

Ozgul, A., Oli, M. K., Armitage, K. B., Blumstein, D. T., and Van Vuren, D. H.
(2009). Influence of local demography on asymptotic and transient dynamics
of a yellow-bellied marmot metapopulation. *The American Naturalist*,
173(4), 517–530.
https://doi.org/10.1086/597225
