---
title: "mpmaggregate: An Introduction to Aggregating Matrix Population Models"
author:
  - name: "Richard A. Hinrichsen"
    orcid: "0000-0003-0761-3005"
output:
  rmarkdown::html_vignette:
    number_sections: false
    mathjax: default
vignette: >
  %\VignetteIndexEntry{mpmaggregate: An Introduction to Aggregating Matrix Population Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
**ORCID:** <https://orcid.org/0000-0003-0761-3005>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Matrix population models (MPMs) are widely used to describe population dynamics, 
but they are often constructed at different levels of age or stage resolution. In 
practice, this can make models difficult to compares side by side, or unnecessarily 
complex for a given analytical goal. The `mpmaggregate` package provides tools for 
reducing the dimensionality of MPMs, allowing finer-scale models 
to be compared with coarser models.

The package supports aggregation of matrix population models under 
two complementary demographic frameworks. The lambda framework focuses on stable 
growth rate, while the R0 framework emphasizes net reproductive rate. 
Within each framework, users may choose between standard aggregation, which maintains 
consistency in the core demographic structure, and elasticity-consistent aggregation, 
which additionally maintains consistency in reproductive values and elasticities.

Two primary functions are provided. The function `mpm_aggregate()` aggregates general 
stage-structured MPMs (Caswell, 2001) to coarser stage classifications, whereas 
leslie_aggregate() is designed specifically for age-structured Leslie matrices 
(Leslie, 1945) and returns a lower-dimensional Leslie matrix that preserves Leslie 
structure.  In both cases, aggregation is accompanied by an effectiveness metric 
that quantifies the degree of agreement between the aggregated and original models. 
Values close to 1 indicate minimal aggregation error and strong agreement, whereas 
smaller values reflect increasing approximation.

Together, these tools make it possible to simplify complex matrix models, compare
models constructed at different resolutions, and explore how demographic conclusions 
depend on the level of aggregation.

## Loading `mpmaggregate`
We begin by loading the `mpmaggregate` package, which provides the functions used 
throughout this vignette for aggregating matrix population models under the \(\lambda\) 
and \(R_0\) demographic frameworks.
```{r}
library(mpmaggregate)
library (expm)
library(knitr)
library(kableExtra)
```

## Aggregating general-to-general MPM

```{r table-preserved, echo=FALSE, results='asis'}

preserved_tbl <- data.frame(
  Framework = c("lambda", "lambda", "R0", "R0"),
  Criterion = c("standard", "elasticity-consistent", "standard", "elasticity-consistent"),
  Preserved = c(
    "<em>λ</em> (stable growth rate)<br>
     Stable stage distribution<br>
     Interstage flow matrix",

    "<em>λ</em> (stable growth rate)<br>
     Stable stage distribution<br>
     Reproductive values<br>
     Elasticities of <em>λ</em><br>
     Generation time (<em>T</em><sub>a</sub>)",

    "<em>R</em><sub>0</sub> (net reproductive rate)<br>
     Cohort stable stage distribution<br>
     Cohort interstage flow matrix",

    "<em>R</em><sub>0</sub> (net reproductive rate)<br>
     Cohort stable stage distribution<br>
     Cohort reproductive values<br>
     Elasticities of <em>R</em><sub>0</sub><br>
     Cohort generation time (<em>T</em><sub>c</sub>)"
  ),
  check.names = FALSE
)

knitr::kable(
  preserved_tbl,
  format = "html",
  escape = FALSE,
  align = c("l", "l", "l"),
  col.names = c("Framework", "Criterion", "Consistent Demographic Quantities"),
caption = paste0(
  "<strong>Table 1.</strong> Demographic quantities that remain consistent under ",
  "<code>mpm_aggregate()</code> in the <em>λ</em> and <em>R</em><sub>0</sub> frameworks ",
  "for both the standard and elasticity-consistent aggregation criteria. ",
  "Under the <em>λ</em> framework, aggregation maintains consistency in stable ",
  "growth properties, whereas under the <em>R</em><sub>0</sub> framework it maintains ",
  "consistency in cohort-based reproductive quantities. ",
  "Interstage flow weights the columns of <strong>A</strong> by their respective ",
  "stable stage densities (Yokomizo et al. 2024). ",
  "Here, “cohort” denotes the <em>R</em><sub>0</sub>-framework analogues of the ",
  "corresponding <em>λ</em>-framework demographic quantities. ",
  "Notice the symmetry between the <em>λ</em> and <em>R</em><sub>0</sub> frameworks."
)
) |>
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "condensed")
  ) |>
  column_spec(3, width = "7cm") |>
  row_spec(0, extra_css = "border-bottom: 2px solid black;") |>
  footnote(
    general_title = "",
    general = paste0(
      "<div style='border-top: 1px solid black; padding-top: 6px;'>",
      "<span style='font-size: 90%;'>",
      "<strong>Note.</strong> <code>mpm_aggregate()</code> retains the projection ",
      "interval of the original matrix population model.",
      "</span></div>"
    ),
    footnote_as_chunk = TRUE,
    escape = FALSE
  )
```


To illustrate aggregation of a general stage-structured MPM, we begin with a simple 
example adapted from Bienvenu et al. (2017). The population projection matrix \( \mathbf{A} \) 
represents a three-stage life cycle, with entries corresponding to survival, transition, 
and reproduction. Although small, this example captures the essential features needed 
to demonstrate aggregation.

We define a grouping of stages using the object `groups`, which specifies how the 
original stages are combined into broader classes. In this example, the first stage 
is left unchanged, while the second and third stages are aggregated into a single 
group. Internally, this grouping is converted to a partition matrix using `mpm_partition()`, 
which is then used to construct the aggregated model.

To illustrate general-to-general MPM aggregation, we begin with a simple 3x3 stage
structured \( \mathbf{A} \) COMPADRE MatrixID = 247940 (Salguero‐Gómez et al., 2015), for a
Marula (*Sclerocarya	birrea*) population, with a seedling, juvenile, and adult stage
(Emanuel et al., 2005). Here, the projection interval is 1 year, and is unchanged by
general-to-general MPM aggregation.

```{r}
A <- matrix(
  c(0.0, 0.0000, 12.46,
    0.4, 0.9160,  0.00,
    0.0, 0.0371,  0.99),
  nrow = 3,
  byrow = TRUE
)
U <- matrix(
  c(0.0, 0.0000,  0.00,
    0.4, 0.9160,  0.00,
    0.0, 0.0371,  0.99),
  nrow = 3,
  byrow = TRUE
)
F <- matrix(
  c(0.0, 0.0000, 12.46,
    0.0, 0.0000,  0.00,
    0.0, 0.0000,  0.00),
  nrow = 3,
  byrow = TRUE
)

#group the vegetative + reproductive stages
#leave the seedling stage alone
groups <- list(
  c(1),
  c(2,3)
)
```

The object `groups` defines how stages are aggregated.
Internally, this is converted to a partition matrix using
`mpm_partition()`.


### Lambda framework: standard aggregation

We first aggregate in the \(\lambda\) framework using **standard aggregation** 
(Hooley, 2000; Salguero-Gomez & Plotkin, 2010). This option is designed to maintain 
consistency in the dominant eigenvalue (\(\lambda\)) and the corresponding stable stage distribution, 
while producing a lower-dimensional matrix  consistent with the grouping defined 
in `groups`. Consistency of interstage flows is also maintained, and this condition 
can be used to derive the standard aggregator (Hinrichsen et al., 2026). 

The call below returns (i) the aggregated projection matrix and (ii) an effectiveness 
value \(\rho^2\), which summarizes how closely the aggregated model reproduces the 
dynamics of the original model. Values of \(\rho^2\) close to 1 indicate near-perfect 
agreement.

```{r}
res_lambda_std <- mpm_aggregate(
  matA = A,
  groups = groups,
  framework = "lambda",
  criterion = "standard"
)

# Aggregated projection matrix
res_lambda_std$matA_agg

# Effectiveness of aggregation
res_lambda_std$effectiveness

# lambda is preserved
spectral_radius(A)
spectral_radius(res_lambda_std$matA_agg)

#stable stage distribution is consistent
P <- mpm_partition(groups=groups,n=3)
w <- stable_stage(A=A)
c(P%*%w)
w_agg <-stable_stage(res_lambda_std$matA_agg)
w_agg

#interstage flows are consistent
P%*%A%*%diag(w)%*%t(P)
res_lambda_std$matA_agg%*%diag(w_agg)
```

### Lambda framework: elasticity-consistent aggregation

Elasticity-consistent aggregation additionally maintains consistency of reproductive 
values. This approach can produce survival probabilities greater than 1 in the aggregated 
matrix (Bienvenu et al. 2017), a behavior that does not occur under standard aggregation. 
Here, consistency of interstage flows is sacrificed in order to maintain consistency 
of the elasticities of (\(\lambda\)) with respect to matrix entries.

```{r}
res_lambda_el <- mpm_aggregate(
  matA = A,
  matF = F,
  matU = U,
  groups = groups,
  framework = "lambda",
  criterion = "elasticity"
)

res_lambda_el$matA_agg
res_lambda_el$effectiveness

#lambda is preserved
spectral_radius(A)
spectral_radius(res_lambda_el$matA_agg)

#stable stage distribution are consistent
P <- mpm_partition(groups=groups,n=3)
w <- stable_stage(A=A)
c(P%*%w)
stable_stage(res_lambda_el$matA_agg)

#reproductive values are consistent
v = reproductive_values(A)
c(solve(P%*%diag(w)%*%t(P))%*%P%*%(w*v))  
reproductive_values(res_lambda_el$matA_agg)

#elasticities are consistent
EA <- elasticity(matA=A,framework="lambda")$elasticity
P%*%EA%*%t(P)
elasticity(matA=res_lambda_el$matA_agg,framework="lambda")$elasticity

#generation time is preserved
generation_time(matF=F,matU=U,framework="lambda")$generation_time
generation_time(matF=res_lambda_el$matF_agg,
                matU=res_lambda_el$matU_agg,
                framework="lambda")$generation_time
```
**Notice** that the [2,1] entry, which is survival probability of the  aggregated matrix 
`res_lambda_el$matA_agg` exceeds 1, which is a known difficulty with the elasticity-consistent 
aggregator (Bienvenu et al. 2017). 

### \(R_0\)framework: decomposition of \( \mathbf{A} \)

We now illustrate aggregation under the \(R_0\) framework, which is based on the
decomposition of the population projection matrix into survival and reproduction
components. Rather than working directly with the full projection matrix \( \mathbf{A} \),
aggregation in this framework is performed using the survival matrix \( \mathbf{U} \) and
the fertility matrix \( \mathbf{F} \).

For this example, we use the same life cycle as above, with \( \mathbf{U} \) describing
survival and transitions among stages and \( \mathbf{F} \) describing reproduction. Together,
these matrices satisfy \( \mathbf{A} = \mathbf{U} + \mathbf{F} \)
.


### \(R_0\) framework: standard aggregation

We first apply standard aggregation in the \(R_0\) framework. In this setting,
aggregation is carried out using the survival and reproduction components of the
model and is designed to ensure consistency of the net reproductive rate (\(R_0\)) and its
associated cohort stable stage distribution. Also consistent is the cohort
interstage flow matrix which weights the columns of \( \mathbf{A} \) by their associated
cohort stable stage densities.


As before, the function returns the aggregated projection matrix and an effectiveness
value that summarizes how closely the aggregated model reproduces the demographic
behavior of the original model.

```{r}
res_R0_std <- mpm_aggregate(
  matU = U,
  matF = F,
  groups = groups,
  framework = "R0",
  criterion = "standard"
)

# Aggregated projection matrix
res_R0_std$matA_agg

# Effectiveness of aggregation
res_R0_std$effectiveness

# R0 is preserved
spectral_radius(F%*%solve(diag(3)-U))
spectral_radius(res_R0_std$matF_agg%*%solve(diag(2)-res_R0_std$matU_agg))

#stable stage distribution is consistent
w <- stable_stage(solve(diag(3)-U)%*%F)
c(P%*%stable_stage(solve(diag(3)-U)%*%F))
w_agg <- stable_stage(solve(diag(2)-res_R0_std$matU_agg)%*%res_R0_std$matF_agg)
w_agg

#cohort interstage flows are consistent
P%*%A%*%diag(w)%*%t(P)
res_R0_std$matA_agg%*%diag(w_agg)
```


### \(R_0\) framework: elasticity-consistent aggregation

Elasticity-consistent aggregation in the \(R_0\) framework additionally ensures
consistent cohort reproductive values and the elasticities of \(R_0\) with respect to matrix
entries. As in the \(\lambda\) framework, this approach can produce survival
probabilities greater than 1 in the aggregated matrix. This behavior reflects the
constraints imposed by elasticity preservation and does not occur under standard
aggregation or in Leslie-to-Leslie aggregation.

```{r}
res_R0_el <- mpm_aggregate(
  matU = U,
  matF = F,
  groups = groups,
  framework = "R0",
  criterion = "elasticity"
)

# Aggregated projection matrix
res_R0_el$matA_agg

# Effectiveness of aggregation
res_R0_el$effectiveness

# R0 is preserved
R0 <- spectral_radius(F%*%solve(diag(3)-U))
R0
spectral_radius(res_R0_el$matF_agg%*%solve(diag(2)-res_R0_el$matU_agg))

#cohort stable stage distribution is consistent
P <- mpm_partition(groups=groups,n=3)
Aref = F + U*R0
w <- stable_stage(A=Aref)
c(P%*%w)
Aref_agg = res_R0_el$matF_agg+ R0*res_R0_el$matU_agg
stable_stage(Aref_agg)

#reproductive values are consistent
v = reproductive_values(Aref)
c(solve(P%*%diag(w)%*%t(P))%*%P%*%(w*v))  
reproductive_values(Aref_agg)

#elasticities of R0 are consistent
#Here elasticities are normalized to sum to 1
EA <- elasticity(matF=F,matU=U,framework="R0")$elasticity
P%*%EA%*%t(P)
elasticity(matF=res_R0_el$matF_agg,matU=res_R0_el$matU_agg,framework="R0")$elasticity

#cohort generation time is preserved
generation_time(matF=F,matU=U,framework="R0")$generation_time
generation_time(matF = res_R0_el$matF_agg, 
                matU = res_R0_el$matU_agg,
                framework="R0")$generation_time
```

## Aggregating Leslie-to-Leslie MPM

```{r table-leslie-preserved, echo=FALSE, results='asis'}
preserved_leslie_tbl <- data.frame(
  Framework = c("lambda", "lambda", "R0", "R0"),
  Criterion = c("standard", "elasticity-consistent", "standard", "elasticity-consistent"),
  Preserved = c(
    "<em>λ</em> (stable growth rate)<br>
     Stable age distribution<br>
     Interstage flow matrix of <strong>A</strong><sup><em>k</em></sup>",

    "<em>λ</em> (stable growth rate)<br>
     Stable age distribution<br>
     Reproductive values<br>
     Elasticities of <em>λ</em><sup><em>k</em></sup> with respect to the entries of <strong>A</strong><sup><em>k</em></sup><br>
     Generation time (<em>T</em><sub>a</sub>) of <strong>A</strong><sup><em>k</em></sup>",

    "<em>R</em><sub>0</sub> (net reproductive rate)<br>
     Cohort stable age distribution<br>
     consistent of <strong>A</strong><sub><em>k</em></sub><sup></sup>",

    "<em>R</em><sub>0</sub> (net reproductive rate)<br>
     Cohort stable age distribution<br>
     Cohort reproductive values<br>
     Elasticities of <em>R</em><sub>0</sub> with respect to <strong>A</strong><sub><em>k</em></sub><sup></sup><br>
     Cohort generation time (<em>T</em><sub>c</sub>) for <strong>A</strong><sub><em>k</em></sub><sup></sup>"
  ),
  check.names = FALSE
)

knitr::kable(
  preserved_leslie_tbl,
  format = "html",
  escape = FALSE,
  align = c("l", "l", "l"),
  col.names = c("Framework", "Criterion", "Consistent Demographic Quantities"),
caption = paste0(
  "<strong>Table 2.</strong> Summary of the demographic properties that remain consistent under ",
  "<code>leslie_aggregate()</code> when reducing a Leslie matrix from dimensionality ",
  "<em>n</em> to <em>m</em>. The quantities the remain consistent depend on both the demographic framework ",
  "(long-term growth under <em>λ</em> or lifetime reproductive output under ",
  "<em>R</em><sub>0</sub>) and the aggregation criterion (standard versus ",
  "elasticity-consistent). The integer <em>k</em> = <em>n</em>/<em>m</em> defines the ",
  "temporal rescaling by aggregating a Leslie model. Quantities in the ",
  "<em>λ</em> framework are defined for <strong>A</strong><sup><em>k</em></sup>, ",
  "while cohort quantities in the <em>R</em><sub>0</sub> framework are defined for ",
  "<strong>A</strong><sub><em>k</em></sub><sup>&dagger;</sup>. Notice the structural symmetry ",
  "between the <em>λ</em> and <em>R</em><sub>0</sub> frameworks."
)
) |>
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "condensed")
  ) |>
  column_spec(3, width = "7cm") |>
  row_spec(0, extra_css = "border-bottom: 2px solid black;") |>
  footnote(
    general_title = "",
    general = paste0(
      "<div style='border-top: 1px solid black; padding-top: 6px;'>",
      "<span style='font-size: 90%;'>",
      "<strong>Note.</strong> Aggregation with <code>leslie_aggregate()</code> ",
      "changes the projection interval from Δt to <em>k</em>Δt.<br>",
      "<sup>&dagger;</sup> <strong>A</strong><sub><em>k</em></sub><sup></sup> denotes the <em>k</em>-step-ahead projection matrix ",
      "in the <em>R</em><sub>0</sub> framework.</span></div>"
    ),
    footnote_as_chunk = TRUE,
    escape = FALSE
  )
```

In some applications, it is desirable to aggregate an age-structured Leslie matrix
while retaining its Leslie form (Hinrichsen 2023; Hinrichsen et al., 2026). The 
function `leslie_aggregate()` is designed for this purpose. It aggregates a Leslie 
matrix with fine-grained age classes into a coarser Leslie matrix of dimensionality
`ngroups`.

To illustrate Leslie-to-Leslie MPM aggregation, we begin with a simple 4x4 Leslie matrix 
\( \mathbf{A} \) COMADRE MatrixID = 248187 (Salguero‐Gómez et al., 2016), for a 
fathead minnow (*Pimephales	promelas*) population (Spromberg & Birge, 2005).

```{r}
A <- rbind(
  c(0.00, 0.00,  6.00, 62.50),
  c(0.50, 0.00,  0.00,  0.00),
  c(0.00, 0.18,  0.00,  0.00),
  c(0.00, 0.00,  0.12,  0.00)
)

FA <- rbind(
  c(0.00, 0.00,  6.00, 62.50),
  c(0.00, 0.00,  0.00,  0.00),
  c(0.00, 0.00,  0.00,  0.00),
  c(0.00, 0.00,  0.00,  0.00)
)

UA <- rbind(
  c(0.00, 0.00,  0.00,  0.00),
  c(0.50, 0.00,  0.00,  0.00),
  c(0.00, 0.18,  0.00,  0.00),
  c(0.00, 0.00,  0.12,  0.00)
)
   
```


### Lambda framework: standard aggregation

We first aggregate the Leslie matrix in the \(\lambda\) framework using standard
aggregation (Hinrichsen, 2023). As in the general-to-general case, this approach ensures
consistency of the stable population growth rate (\(\lambda\)) and the stable age distribution,
while producing a Leslie matrix of reduced dimensionality. 
The interstage flow of the aggregated matrix is consistent with that of
\( \mathbf{A}^{k} \).

```{r}
res_leslie_lambda_std <- leslie_aggregate(
  matA = A,
  ngroups = 2,
  framework = "lambda",
  criterion = "standard"
)

P <- rbind(c(1,1,0,0),
           c(0,0,1,1))

n <- 4 #dimensionality of A
m <- 2 #dimensionality of aggregated form of A
k <- n/m

# Aggregated Leslie matrix
res_leslie_lambda_std$matAk_agg

# Effectiveness of aggregation
res_leslie_lambda_std$effectiveness

# lambda is preserved
spectral_radius(A)^k
spectral_radius(res_leslie_lambda_std$matAk_agg)

#stable stage distribution is consistent
w <- leslie_stable_age(A=A)
c(P%*%w)
w_agg <-leslie_stable_age(res_leslie_lambda_std$matAk_agg)
w_agg

#interstage flows of L^k are consistent
P%*%(A%^%k)%*%diag(w)%*%t(P)
res_leslie_lambda_std$matAk_agg%*%diag(w_agg)
```

**Note.** In this case, the dimensionality of the aggregated matrix (\(m = 2\))
divides the dimensionality of \( \mathbf{A} \) (\(n = 4\)) evenly. More generally, when these
dimensionalities are not compatible, a disaggregation step is used internally
to expand \( \mathbf{L} \) to a compatible dimensionality prior to aggregation 
(Hinrichsen, 2023).

### Lambda framework: elasticity-consistent aggregation

We next apply elasticity-consistent aggregation to the Leslie matrix in the
\(\lambda\) framework (Hinrichsen et al., 2026). In addition to yielding consistent stable population growth
rate and stable age distribution, this approach also yields consistent reproductive values and
the elasticities of \(\lambda^{k}\) with respect to the entries entries of \( \mathbf{A}^{k} \).
Unlike elasticity-consistent aggregation for general stage-structured MPMs, Leslie-to-Leslie
elasticity-consistent aggregation does not produce survival probabilities greater
than 1.

```{r}
res_leslie_lambda_el <- leslie_aggregate(
  matA = A,
  ngroups = 2,
  framework = "lambda",
  criterion = "elasticity"
)

# Aggregated Leslie matrix
res_leslie_lambda_el$matAk_agg

# Effectiveness of aggregation
res_leslie_lambda_el$effectiveness

#get aggregated F and U
F_agg <-matrix(0,2,2)
F_agg[1,] <- res_leslie_lambda_el$matAk_agg[1,]
U_agg <- res_leslie_lambda_el$matAk_agg
U_agg[1,] <-0

# lambda is preserved
spectral_radius(A)^k
spectral_radius(res_leslie_lambda_el$matAk_agg)

#stable stage distribution is consistent
w <- leslie_stable_age(A=A)
c(P%*%w)
w_agg <-leslie_stable_age(res_leslie_lambda_el$matAk_agg)
w_agg

#reproductive values are preserved
v <- leslie_reproductive_values(A=A)
c(solve(P%*%diag(w)%*%t(P))%*%P%*%(v*w))
leslie_reproductive_values(A=res_leslie_lambda_el$matAk_agg)

#elasticities of lambda^k  are consistent
EA <- elasticity(A%^%k,framework="lambda")$elasticity
P%*%EA%*%t(P)
elasticity(res_leslie_lambda_el$matAk_agg,framework="lambda")$elasticity

#generation times are consistent between L^k and aggregated matrix
U_k <- UA%^%k
F_k <- A%^%k - U_k
generation_time(matF=F_k,matU=U_k,framework="lambda")$generation_time
generation_time(matF=F_agg,matU=U_agg,framework="lambda")$generation_time
```
---
### \(R_0\) framework

Leslie-to-Leslie aggregation can also be carried out under the **\(R_0\) framework**,
emphasizing net reproductive rate rather than stable population
growth. As in the general-to-general case, aggregation in this framework is based
on the underlying survival and fertility components of the Leslie matrix,
but the result again retains Leslie form.

### \(R_0\) framework: standard aggregation

We first apply **standard aggregation** in the \(R_0\) framework. This approach yields
consistent net reproductive rate (\(R_0\)) and the cohort stable age distribution, while
producing a coarser Leslie matrix compatible with the specified number of age groups.

```{r}
res_leslie_R0_std <- leslie_aggregate(
  matA = A,
  ngroups = 2,
  framework = "R0",
  criterion = "standard"
)

# Aggregated Leslie matrix
res_leslie_R0_std$matAk_agg

# Effectiveness of aggregation
res_leslie_R0_std$effectiveness

#get F and U for aggregated matrix
FA_agg <- matrix(0,2,2)
FA_agg[1,] <- res_leslie_R0_std$matAk_agg[1,]
UA_agg <- res_leslie_R0_std$matAk_agg
UA_agg[1,] <- 0

# R0 is preserved
R0 <- spectral_radius(FA%*%solve(diag(4)-UA))
R0 
spectral_radius(FA_agg%*%solve(diag(2)-UA_agg))


#cohort stable stage distribution is consistent
A_ref <- (FA + UA*R0)/R0
A_ref_agg <- (FA_agg + UA_agg*R0)/R0
w <- leslie_stable_age(A=A_ref)
c(P%*%w)
w_agg <-leslie_stable_age(A_ref_agg)
w_agg

#cohort interstage flows of A_k (K-step ahead version of A in R0 framework) are consistent
U_k <- UA%^%k
F_k <- R0*(A_ref)%^%k-R0*U_k
A_k <- F_k + U_k
P%*%A_k%*%diag(w)%*%t(P)
(FA_agg + UA_agg)%*%diag(w_agg)
```

### \(R_0\) framework: elasticity-consistent aggregation

We now apply **elasticity-consistent aggregation** in the \(R_0\) framework. In addition
to providing consistent net reproductive rate (\(R_0\)) and the cohort stable age distribution,
this approach provides consistent cohort reproductive values and the elasticities of \(R_0\)
with respect to the entries of \( \mathbf{A}_k^{*} \), the \(k\)-step–ahead version of \( \mathbf{A} \), where
\(k = n/m\), \(n\) is the dimensionality of the original model, and \(m\) is the
dimensionality of the aggregated model. As in the \(\lambda\) framework, the Leslie
structure is maintained and survival probabilities cannot exceed 1.

```{r}
res_leslie_R0_el <- leslie_aggregate(
  matA = A,
  ngroups = 2,
  framework = "R0",
  criterion = "elasticity"
)

# Aggregated Leslie matrix
res_leslie_R0_el$matAk_agg

# Effectiveness of aggregation
res_leslie_R0_el$effectiveness

#get F and U for aggregated matrix
FA_agg <- matrix(0,2,2)
FA_agg[1,] <- res_leslie_R0_el$matAk_agg[1,]
UA_agg <- res_leslie_R0_el$matAk_agg
UA_agg[1,] <- 0

# R0 is preserved
R0 <- spectral_radius(FA%*%solve(diag(4)-UA))
R0 
spectral_radius(FA_agg%*%solve(diag(2)-UA_agg))


#cohort stable stage distribution is consistent
A_ref <- (FA + UA*R0)/R0
A_ref_agg <- (FA_agg + UA_agg*R0)/R0
w <- leslie_stable_age(A=A_ref)
c(P%*%w)
w_agg <-leslie_stable_age(A_ref_agg)
w_agg

#cohort reproductive values are consistent
v <- leslie_reproductive_values(A_ref)
c(solve(P%*%diag(w)%*%t(P))%*%P%*%(v*w))
leslie_reproductive_values(A_ref_agg)

#elasticities of R0 are consistent
#these elasticities are normalized to sum to 1
U_k <- UA%^%k
F_k <- R0*(A_ref)%^%k-R0*U_k
A_k <- F_k + U_k
EA <- elasticity(matF=F_k,matU=U_k,framework="R0")$elasticity
P%*%EA%*%t(P)
elasticity(matF=FA_agg,matU=UA_agg,framework="R0")$elasticity

#cohort generation times are consistent between A_k an aggregated matrix
generation_time(matF=F_k, matU = U_k, framework="R0")$generation_time
generation_time(matF=FA_agg,matU=UA_agg, framework="R0")$generation_time
```



## Interpretation

- The **lambda framework** ensures consistency of stable growth and stable structure.
- The **R0 framework** ensures consistency of generation-based reproductive dynamics.
- **Standard aggregation** ensures consistency of the core demographic quantity associated
  with the chosen framework (either \(\lambda\) or \(R_0\)).
- **Elasticity-consistent aggregation** additionally ensures consistency of reproductive
  values, associated elasticities, and generation times within their
  respective frameworks.


## References

Bienvenu, F., Akçay, E., Legendre, S. and McCandlish, D.M. 2017. The genealogical
decomposition of a matrix population model with applications to the aggregation
of stages. Theoretical Population Biology, 115, 69-80. 
https://doi.org/10.1016/j.tpb.2017.04.002

Caswell, H. 2001. Matrix Population Models: Construction, Analysis, and Interpretation. 
2nd edition. Sinauer Associates, Sunderland, MA. 

Emanuel, P.L., Shackleton, C.M. and Baxter, J.S., 2005. Modelling the sustainable 
harvest of Sclerocarya birrea subsp. caffra fruits in the South African lowveld. 
Forest Ecology and Management, 214(1-3), 91-103. 
https://doi.org/10.1016/j.foreco.2005.03.066

Hinrichsen, R. A. 2023. Aggregation of Leslie matrix models with application to 
ten diverse animal species. Population Ecology, 65(3), 146–166. 
https://doi.org/10.1002/1438-390X.12149

Hinrichsen, R. A., Yokomizo, H., & Salguero-Gómez, R. 2026. From theory to
application: Elasticity-consistent aggregation of Leslie matrix population
models for comparative demography. bioRxiv, preprint.
https://doi.org/10.64898/2026.02.04.703802

Hooley, D. E. 2000. Collapsed matrices with (almost) the same eigenstuff. The 
College Mathematics Journal, 31(4), 297–299. https://doi.org/10.1080/07468342.2000.11974162

Leslie, P. H. 1945. On the use of matrices in certain population mathematics. Biometrika, 
33(3), 183–212. https://doi.org/10.2307/2332297


Salguero-Gómez, R., Jones, O. R., Archer, C. R., Buckley, Y. M., Che-Castaldo, J., 
Caswell, H., Hodgson, D., Scheuerlein, A., Conde, D. A., Brinks, E., de Buhr, H., 
Farack, C., Gottschalk, F., Hartmann, A., Henning, A., Hoppe, G., Römer, G., 
Runge, J., Ruoff, T., Wille, J., Zeh, S., Davison, R., Vieregg, D., Baudisch, A., 
Altwegg, R., Colchero, F., Dong, M., de Kroon, H., Lebreton, J. -D., Metcalf, C. J. E., 
Neel, M. M., Parker, I. M., Takada, T., Valverde, T., Vélez-Espino, L. A., Wardle, G. M., 
Franco, M., & Vaupel, J. W. (2015). The COMPADRE Plant Matrix Database: an open 
online repository for plant demography. Journal of Ecology, 103, 202–218. 
https://doi.org/10.1111/1365-2745.12334

Salguero‐Gómez, R., Jones, O. R., Archer, C. R., Bein, C., de Buhr, H., Farack, C., 
Gottschalk, F., Hartmann, A., Henning, A., Hoppe, G., Römer, G., Ruoff, T., Sommer, V., 
Wille, J., Voigt, J., Zeh, S., Vieregg, D., Buckley, Y. M., Che‐Castaldo, J., Hodgson, D., 
Scheuerlein, A., Caswell, H., & Vaupel, J. W. 2016. COMADRE: a global database of animal 
demography. Journal of Animal Ecology, 85, 371–384. https://doi.org/10.1111/1365-2656.12482

Salguero-Gomez, R. & Plotkin, J. B. 2010. Matrix dimensions bias
demographic inferences: implications for comparative plant demography. The
American Naturalist 176, 710-722. https://doi.org/10.1086/657044

Spromberg, J.A. and Birge, W.J., 2005. Modeling the effects of chronic toxicity 
on fish populations: The influence of life‐history strategies. Environmental 
Toxicology and Chemistry, 24(6), 1532-1540. https://doi.org/10.1897/04-160.1

Yokomizo, H., Fukaya, K., Lambrinos, J.G., Kawai, Y. and Takada, T., 2024. 
Interstage flow matrices: Population statistic derived from matrix population models. 
Journal of Ecology, 112(6), 1326-1338.https://doi.org/10.1111/1365-2745.14303

